---
title: "Do LGBTQ+ YP have different aspirations for their future?"
date: '2023-06-16'
categories: ['lgbtq+', 'aspirations', 'school leavers']
format: html
editor: visual
---
  
```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(purrr)
library(tidyverse)
library(knitr)
library(simplermarkdown)

box::use(../../../phei_functions/hes)

suppress_everything <- function(df) {
  
  df %>% mutate(across(where(is.numeric), ~ if_else(.x <= 7 & .x != 0, NA, .x) %>% `/`(5) %>% round() %>% `*`(5)))
  
}

```

```{r}
yphws_data_private <- readRDS("~/projects/yphws_analysis/data/private/yphws_data_private.Rds")

count <- yphws_data_private$sexuality %>% table() %>% as_tibble() %>% rename("answer" = ".", "number" = "n")

count <- suppress_everything(count)

total_students <- nrow(yphws_data_private)

count <- count %>% mutate("%" = round((number/total_students)*100, 1))
```

#### Abstract

Young LGBTQ+ students face different lived experiences than those from non-care homes. This could influence their decision-making around aspirations after leaving school, such as their career, education, or training. This brief analysis seeks to ascertain whether a student's sexual orientation has a relationship with their aspirations after school.

```{r}
#| output: asis
print(kable(count))
```

: Summary count of students on question asking about sexual orientation in YPHWS {#tbl-summary-count}

